---
title: "Differential Abundance Testing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We'll look into using DESeq2 to test for differential abundance in a phyloseq object, as introduced in [this vignette](https://bioconductor.org/packages/devel/bioc/vignettes/phyloseq/inst/doc/phyloseq-mixture-models.html)

The data are from a public dataset on [colorectal
cancer](http://genome.cshlp.org/content/22/2/292.long).

First we download and load the data.

```{r libraries}
library("phyloseq")
filepath = system.file("extdata", "study_1457_split_library_seqs_and_mapping.zip", package="phyloseq")
kostic = microbio_me_qiime(filepath)
head(sample_data(kostic)$DIAGNOSIS, 10)
```

We're interested in whether some species are more or less abundant in healthy
vs. tumor samples.

```{r subset}
kostic <- subset_samples(kostic, DIAGNOSIS != "None")
kostic <- prune_samples(sample_sums(kostic) > 500, kostic)
kostic

kostic_dds <- phyloseq_to_deseq2(kostic, ~ DIAGNOSIS)
kostic_dds
```

This single function does all the estimation and inference. Note that it
transforms the `kostic_dds` object in place.

```{r}
library("DESeq2")
kostic_dds <- DESeq(kostic_dds, fitType = "local", sfType = "poscounts")
```

Can you inspect the results?

* Produce an MA plot of species-wise effects.
* Produce a plot of the estimated dispersions. Which dispersions are not shrunk?
* Make a histogram of the estimated size factors.
* Which species are most significantly associated with colorectal cancer?
* Experiment with different dispersion and size factor estimation methods. How
  sensitive are the results?
  
Some functions that might be useful, `results()`, `plotDispEsts()`, `plotMA()`,
`colData()`, `design()`....

Bonus questions,
* Maybe we want to control for radiation therapy when testing for the
  association. Figure out how to modify the design matrix to accomodate this.
* Read about "Heavy-tailed prior distributions for sequence count data: removing
  the noise and preserving large differences" and how to implement that approach
  in our kostic analysis.
* Read about independent hypothesis weighting as an alternative to simple
  Benjamini-Hochberg, and how to impelment that using DESeq2.
